Supporting Information for: A Standardized Effect Size for Evaluating and Comparing the Strength of Phylogenetic Signal

### 2020-09-17 Manuscript/PNAS-Adams-Submitted.Rmd
new Z12 compare

  Data deposition: Data for the empirical example may be found on DRYAD: doi:10.5061/dryad.b554m44 and doi:10.5061/dryad.59zw3r23m. R-scripts for simulation tests are found on Github: XXX. Computer code for implementing the two-sample comparison of effect sizes is found in geomorph: https://cran.r-project.org/web/packages/geomorph/index.html \hfill\break

@@

where $K_1$, $K_2$, $\hat\mu_{K_1}$, $\hat\mu_{K_2}$, $\hat\sigma_{K_1}$, and $\hat\sigma_{K_2}$ are as defined above. Estimates of significance of $\hat{Z}_{12}$ may be obtained from a standard normal distribution, or permutation. Typically, $\hat{Z}_{12}$ is considered a two-tailed test, however directional (one-tailed) tests may be specified should the empirical situation require it [@AdamsCollyer2016; @AdamsCollyer2019b]. Simulations reveal that tests based on $\hat{Z}_{12}$ have appropriate type I error rates (~0.05) and reasonable model misspecification rates (~7-12%: Supporting Information).

>>>>

  Data deposition: Data for the empirical example may be found on DRYAD: doi:10.5061/dryad.b554m44 and doi:10.5061/dryad.59zw3r23m. R-scripts for simulation tests are found at: https://github.com/deanadams/PhySigCompareZ. Computer code for implementing the two-sample comparison of effect sizes is found in geomorph: https://cran.r-project.org/web/packages/geomorph/index.html \hfill\break

@@

where $K_1$, $K_2$, $\hat\mu_{K_1}$, $\hat\mu_{K_2}$, $\hat\sigma_{K_1}$, and $\hat\sigma_{K_2}$ are as defined above. Estimates of significance of $\hat{Z}_{12}$ may be obtained from a standard normal distribution, or permutation. Typically, $\hat{Z}_{12}$ is considered a two-tailed test, however directional (one-tailed) tests may be specified should the empirical situation require it [@AdamsCollyer2016; @AdamsCollyer2019b]. Simulations reveal that tests based on $\hat{Z}_{12}$ have appropriate type I error and model misspecification rates  of approximately 5% (Supporting Information).

### 2020-09-17 Manuscript/PNAS-Adams-Submitted.tex
new Z12 compare

Data deposition: Data for the empirical example may be found on DRYAD:
\url{doi:10.5061/dryad.b554m44} and \url{doi:10.5061/dryad.59zw3r23m}.
R-scripts for simulation tests are found on Github: XXX. Computer code
for implementing the two-sample comparison of effect sizes is found in
geomorph:
\url{https://cran.r-project.org/web/packages/geomorph/index.html}
\hfill\break

@@

where \(K_1\), \(K_2\), \(\hat\mu_{K_1}\), \(\hat\mu_{K_2}\),
\(\hat\sigma_{K_1}\), and \(\hat\sigma_{K_2}\) are as defined above.
Estimates of significance of \(\hat{Z}_{12}\) may be obtained from a
standard normal distribution, or permutation. Typically,
\(\hat{Z}_{12}\) is considered a two-tailed test, however directional
(one-tailed) tests may be specified should the empirical situation
require it (36, 38). Simulations reveal that tests based on
\(\hat{Z}_{12}\) have appropriate type I error rates
(\textasciitilde0.05) and reasonable model misspecification rates
(\textasciitilde7-12\%: Supporting Information).

>>>>

Data deposition: Data for the empirical example may be found on DRYAD:
\url{doi:10.5061/dryad.b554m44} and \url{doi:10.5061/dryad.59zw3r23m}.
R-scripts for simulation tests are found at:
\url{https://github.com/deanadams/PhySigCompareZ}. Computer code for
implementing the two-sample comparison of effect sizes is found in
geomorph:
\url{https://cran.r-project.org/web/packages/geomorph/index.html}
\hfill\break

@@

where \(K_1\), \(K_2\), \(\hat\mu_{K_1}\), \(\hat\mu_{K_2}\),
\(\hat\sigma_{K_1}\), and \(\hat\sigma_{K_2}\) are as defined above.
Estimates of significance of \(\hat{Z}_{12}\) may be obtained from a
standard normal distribution, or permutation. Typically,
\(\hat{Z}_{12}\) is considered a two-tailed test, however directional
(one-tailed) tests may be specified should the empirical situation
require it (36, 38). Simulations reveal that tests based on
\(\hat{Z}_{12}\) have appropriate type I error and model
misspecification rates of approximately 5\% (Supporting Information).

### 2020-09-17 Manuscript/SupportingInfo.Rmd
new Z12 compare

Here we examine the statistical properties of tests for comparing the relative strength of phylogenetic signal, using $\hat{Z}_{12}$. Simulations were performed using pure-birth trees of different sizes ($n=2^5, 2^6, \cdots, 2^{10}$), and with differing levels of phylogenetic signal ($\lambda=0.0, 0.5, \cdots, 1.0$). For each combination of $n$ and $\lambda$, 50 traits were simulated on the phylogeny under a Brownian motion model of evolution. Next, the phylogenetic signal between traits was compared in pairwise fashion for all $50*49/2 = 1225$ combinations, using $\hat{Z}_{12}$. The proportion of significant results provided an estimate of the type I error or model misspecification. Specifically, type I error was evaluated when traits were simulated using no input phylogenetic signal (i.e., $\lambda = 0$). Likewise, model misspecification was evaluated for data simulated with some known, but equal, level of phylogenetic signal (i.e., $\lambda > 0$). \hfill\break

**Results:** Tests revealed that across simulation conditions, the average type I error of $\hat{Z}_{12}$ was approximately 0.05 (Fig. S21). Levels of misspecification were slightly higher, and ranged on average between 7% when comparing traits generated on the same phylogeny (Fig. S21 black line) to ~12% for comparisons of traits from different trees. Both results demonstrate that comparisons of the strength of phylogenetic signal using $\hat{Z}_{12}$ do not display high levels of false positives, and are thus appropriate for comparing the degree of phylogenetic signal across traits.

>>>>

Here we examine the statistical properties of tests for comparing the relative strength of phylogenetic signal, using $\hat{Z}_{12}$. Simulations were performed using pure-birth trees of different sizes ($n=2^5, 2^6, \cdots, 2^{10}$), and with differing levels of phylogenetic signal ($\lambda=0.0, 0.5, \cdots, 1.0$). For each combination of $n$ and $\lambda$, 20 pure-birth trees were generated and 50 traits were simulated on each phylogeny under a Brownian motion model of evolution. Next, the phylogenetic signal between traits was compared in pairwise fashion for all $50*49/2 = 1225$ combinations, using $\hat{Z}_{12}$. The proportion of significant results provided an estimate of the type I error or model misspecification. Specifically, type I error was evaluated when traits were simulated using no input phylogenetic signal (i.e., $\lambda = 0$). Likewise, model misspecification was evaluated for data simulated with some known, but equal, level of phylogenetic signal (i.e., $\lambda > 0$). \hfill\break

**Results:** Tests revealed that across simulation conditions, the average type I error of $\hat{Z}_{12}$ was approximately 0.05 (Fig. S21). Likewise, levels of misspecification were also low, and were at or below the nominal 5% level (Fig. S21 black line). Both results demonstrate that comparisons of the strength of phylogenetic signal using $\hat{Z}_{12}$ do not display high levels of false positives, and are thus appropriate for comparing the degree of phylogenetic signal across traits.

### 2020-09-17 Manuscript/SupportingInfo.tex
new Z12 compare

Here we examine the statistical properties of tests for comparing the
relative strength of phylogenetic signal, using \(\hat{Z}_{12}\).
Simulations were performed using pure-birth trees of different sizes
(\(n=2^5, 2^6, \cdots, 2^{10}\)), and with differing levels of
phylogenetic signal (\(\lambda=0.0, 0.5, \cdots, 1.0\)). For each
combination of \(n\) and \(\lambda\), 101 traits were simulated on the
phylogeny under a Brownian motion model of evolution. Next, the
phylogenetic signal of trait 1 was compared to the signal in the
remaining 100 traits in pairwise fashion, using \(\hat{Z}_{12}\). The
proportion of significant results provided an estimate of the type I
error or model misspecification. Specifically, type I error was
evaluated when traits were simulated using no input phylogenetic signal
(i.e., \(\lambda = 0\)). Likewise, model misspecification was evaluated
for data simulated with some known, but equal, level of phylogenetic
signal (i.e., \(\lambda > 0\)). \hfill\break

\textbf{Results:} Tests revealed that across simulation conditions, the
average type I error of \(\hat{Z}_{12}\) was approximately 0.05 (Fig.
S21). Levels of misspecification were slightly higher, and ranged on
average between 7\% (when comparing traits generated on the same
phylogeny: Fig. S21) to \textasciitilde12\% (for comparisons of traits
from different trees). Both results demonstrate that comparisons of the
strength of phylogenetic signal using \(\hat{Z}_{12}\) do not display
high levels of false positives, and are thus appropriate for comparing
the degree of phylogenetic signal across traits.

\includegraphics[width=28.93in,height=0.5\textheight]{fig.S21}

\textbf{Figure S21}. Type I error rates and model misspecification for
\(\hat{Z}_{12}\). Type I error is found on the far-left of the plot
(\(\lambda = 0\)).

>>>>

Here we examine the statistical properties of tests for comparing the
relative strength of phylogenetic signal, using \(\hat{Z}_{12}\).
Simulations were performed using pure-birth trees of different sizes
(\(n=2^5, 2^6, \cdots, 2^{10}\)), and with differing levels of
phylogenetic signal (\(\lambda=0.0, 0.5, \cdots, 1.0\)). For each
combination of \(n\) and \(\lambda\), 20 pure-birth trees were generated
and 50 traits were simulated on each phylogeny under a Brownian motion
model of evolution. Next, the phylogenetic signal between traits was
compared in pairwise fashion for all \(50*49/2 = 1225\) combinations,
using \(\hat{Z}_{12}\). The proportion of significant results provided
an estimate of the type I error or model misspecification. Specifically,
type I error was evaluated when traits were simulated using no input
phylogenetic signal (i.e., \(\lambda = 0\)). Likewise, model
misspecification was evaluated for data simulated with some known, but
equal, level of phylogenetic signal (i.e., \(\lambda > 0\)).
\hfill\break

\textbf{Results:} Tests revealed that across simulation conditions, the
average type I error of \(\hat{Z}_{12}\) was approximately 0.05 (Fig.
S21). Likewise, levels of misspecification were also low, and were at or
below the nominal 5\% level (Fig. S21 black line). Both results
demonstrate that comparisons of the strength of phylogenetic signal
using \(\hat{Z}_{12}\) do not display high levels of false positives,
and are thus appropriate for comparing the degree of phylogenetic signal
across traits.

\includegraphics[width=6.67in,height=0.5\textheight]{fig.S21}

\textbf{Figure S21}. Type I error rates and model misspecification for
\(\hat{Z}_{12}\). Type I error is found on the far-left of the plot
(\(\lambda = 0\)). The solid black line represents the average type I
error and false discovery rate across all levels of input signal.
